---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)

library(hospEpi)
```

# hospEpi

<!-- badges: start -->
[![R-CMD-check](https://github.com/npeters1322/hospEpi/workflows/R-CMD-check/badge.svg)](https://github.com/npeters1322/hospEpi/actions)
<!-- badges: end -->

The goal of `hospEpi` is to provide useful functions for epidemiological analyses in a hospital setting. Functions so far deal with summarizing and plotting patient location history data and disease-exposure (or risk factor) data.

## Installation

You can install the development version of `hospEpi` like so:

``` r
devtools::install_github("npeters1322/hospEpi", build_vignettes = TRUE)
```

Using `build_vignettes = TRUE` makes sure the package vignette is built when you install the package, which might be useful if you want to explore the vignette.

## Example

After loading the package, say you have some disease-exposure data you want to plot and/or summarize. An example dataset comes with the package and gives a simple idea of what your data should look like (at least for the most part) before using any functions.

```{r}
#use the example disease-exposure dataset from the package
de_data <- disease_expose_data
head(de_data)
```

To work with my plot and summary functions, you'll want a dataset that only contains 0s and 1s in each variable, so you should clean your data using another function in this package if your data is not like that.

```{r}
#clean the disease-exposure data
cleaned_de_data <- clean_disease_expose(data = de_data, disease = "disease", noDisease = "No", exposures = c(3,4,5))
head(cleaned_de_data)
```

As you can see above, all columns are now binary. If you look back at the example data before it was cleaned, you'll see that for the first person, exposure3 was "child." Now, in the cleaned dataset, there is a binary column for `exposure3_child`, and the first row is marked as 1 because that person was marked as a child in the original dataset.  

With cleaned data, we can now make an object of class `disease_expose`. There are specific summary and plot functions in my package to work with objects of that class, so if you want to use those, you must create an object like so:

```{r eval = FALSE}
de_object <- disease_expose(cleaned_de_data)
class(de_object)
```
```{r echo = FALSE}
de_object <- new_disease_expose(x = cleaned_de_data, disease = 1, exposures = 2:8)
class(de_object)
```

Calling the `disease_expose` function will pull up a Shiny gadget, which will allow you to choose your disease column and any exposure columns you want to bring along from your data, and then it will create an object of class `disease_expose` if you receive no errors. If a Shiny gadget is not your thing, you can also manually do everything (choose your disease column and exposure columns by typing them into the function) by using the `new_disease_expose` function.  

Now that you have a `disease_expose` object, you can plot it by simply calling the plot function with the necessary arguments. The only required argument is the `disease_expose` object, but you can also add additional arguments that will be passed to the `geom_bar` function. For example:

```{r}
plot(de_object, position = 'dodge')
```

Now, you can see grouped bar charts of your disease-exposure combinations to see if there is anything interesting in the data.  

You can also summarize your data like so:

```{r}
de_summ <- summary(de_object)
de_summ
```

As you can see, the function outputs several different statistics for each disease-exposure combination and saves them in a `data.frame`, which would allow for easy extraction if wanted.  

There are also multiple functions for patient location history data, and you can find out how to use those by reading the package vignette or the help documentation for the functions.
