---
title: "Introduction to hospEpi"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to hospEpi}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

After installing, you can load the library like this:

```{r setup}
library(hospEpi)
```

## Background

This R package, `hospEpi`, is designed to provide useful functions for epidemiological analyses in a hospital setting. I was motivated to create this package because of my work as an intern for the Indiana University Health Infection Prevention team, as their main analyst, my supervisor, often does work as a hospital epidemiologist of sorts and uses R to do some of his analyses. Some examples of what my package does are that it allows people to easily study patient location history/network data and also disease-exposure/disease-risk factor data. Currently, there are a couple of other R packages that provide tools for epidemiological analyses, like `epiR` and `epitools`. However, my package works a little differently, as it focuses more on applying tools for analyses in a hospital setting. It also addresses a few challenges that make it stand out. Those are:

* It provides multiple plot functions to quickly visualize some of your data. In other packages for epidemiology, plotting seems to mostly be up to the user to figure out. However, in my package, multiple plot functions are provided to visualize the data. For example, you can visualize the network of patient location history data or see how many of those with a certain exposure/risk factor developed a disease. This functionality sets my package apart from others.
* It provides multiple cleaning functions, which can help you get your data in the correct format before using the rest of the functions. Oftentimes, packages are made to work with data that is already in a specific format. In the best case scenario, data coming into my functions would also be clean already, but that might not be the case. With that in mind, my package provides a couple of functions to help clean your data. For example, your disease-exposure data might have columns that are not binary (made up of 0s and 1s), which is necessary for multiple functions in my package. Rather than making you take the time to fix all of that on your own, one of my cleaning functions can help to get your data in the correct format. Another cleaning function is also provided for another use, which will be seen later. Providing cleaning functions sets my package apart from others, as it does not make the user do all the work to clean their data.
* It can easily assist in analyzing multiple exposures/risk factors against a disease, and can quickly report multiple statistics for all of those combinations. In a hospital, you might often look at multiple exposures/risk factors against a disease, like having diabetes, having heart disease, or having surgery during a stay against developing a hospital-acquired disease. Rather than only allowing one to look at one disease-exposure combination at a time, my package has built-in functions that let one study all combinations at once and get many statistics all at once. As a result, you will be able to compare all your exposures against each other to see which seem the most associated with the disease/outcome. 

As seen, while there are other packages for epidemiological analyses, my package adds other functionality that would be helpful. While initially created for working with hospital data, it could easily be used for other epidemiological analyses. It is not an all-encompassing package, so other epidemiology packages should be checked out, but my package does provide a couple useful tools that are not provided elsewhere.

## How to Use hospEpi

As of now, there are two main use cases for this package: analyzing patient location/hospital network data and analyzing disease-exposure/disease-risk factor data. Below, I will walk you through the process of doing each.

#### Working With Patient Location/Hospital Network Data

A good way to start analyzing hospital network data using this package is to look at the example dataset I have created and to go through the examples using it to see if you need to manipulate your data before using the functions. The example data can be loaded in like so:

```{r}
hn_data <- hosp_network_data
head(hn_data)
```

To use many of the functions, the data needs to have columns corresponding to the patient's starting room and next room or starting unit and next unit. Looking at the data above, we do not have those columns yet, so I will use my cleaning function to get everything in the right format.

```{r}
cleaned_hn_data <- clean_hosp_network(data = hn_data, startDate = BeginDate, 
                                      endDate = EndDate, unitName = UnitName, 
                                      roomNum = RoomNumber)
head(cleaned_hn_data)
```

Now that we have clean data, we can move on to creating an object of class `hosp_network`, which will allow the plotting and summary functions to be used. To create the object, you will use the below function.

```{r}
hn_object <- hosp_network(x = cleaned_hn_data, fromUnit = UnitName, toUnit = 
                            next_unit, fromRoom = RoomNumber, toRoom = next_room)
class(hn_object)
```

If you do not have one of the two types of data, room or unit, then you can exclude it like so, as to use the rest of the functions, you just need to make sure to have at least one of the two:

```{r}
hn_object2 <- hosp_network(x = cleaned_hn_data, fromUnit = UnitName, toUnit = next_unit)
```

Now, you can move on to plotting the data. This package creates network graphs using the `igraph` package, which, for example, would allow you to see the network of your patients and how they have moved throughout the hospital. Two examples are:

```{r fig.show = 'hold'}
plot(hn_object, by = "room", type = "simple")
plot(hn_object, by = "unit", type = "hub score", vertex.color = "red", vertex.shape = "square")
```

In the example on the left, the data is plotted by the room data and all points are the same size. In the example on the right, the specifications were changed to using unit data and making the size of the points be based off the hub score (statistic related to graphs, search if wanting to know more). I also added more arguments, which are arguments that can be passed to `plot.igraph`.  

Plotting this type of data could be helpful if you want to visualize how patients have moved throughout the hospital. Perhaps patients have recently been contracting a specific disease in the hospital, and you want to find if there are any common rooms or units among them. If that is the case, these plots can help to visualize that.  

You can also summarize your data.

```{r warning = FALSE}
hn_summ <- summary(hn_object, by = "room")
hn_summ[4:6]
```

This provides many different statistics related to network graphs, which might be helpful if you are interested in knowing more about your network. You can also change it to work with your unit data, and you can also add more arguments that can be passed to `hub_score` and `authority_score`.

```{r warning = FALSE}
hn_summ2 <- summary(hn_object, by = "unit", scale = FALSE)
hn_summ2[4:6]
```

While network statistics might not be useful to everyone, they can provide numerical information about your network.  

Overall, this half of the package provides useful functions for working with patient location history/patient network data.

#### Working With Disease-Exposure/Disease-Risk Factor Data

To begin analyzing disease-exposure/disease-risk factor data, it might be helpful to look at the example dataset that comes with the package and go through the below examples with it. The dataset can be loaded by doing the following:

```{r}
de_data <- disease_expose_data
head(de_data)
```

To use most of my functions, your data must be made of binary columns (0s and 1s). As seen above, the data does not follow that right now, which is why I provided a cleaner function to help, which might also be helpful for you.

```{r}
cleaned_de_data <- clean_disease_expose(data = de_data, disease = "disease", 
                                        noDisease = "No", 
                                        exposures = c("exposure1", "exposure2", "exposure3"))
head(cleaned_de_data)
```

As seen above, all columns are now binary variables, and we are ready to move on to the rest of my functions. Note that although all columns are binary, you might not want every column created by the cleaner function. For example, you might only want one of the two `exposure2_...` columns, as they are just opposites of each other. For now, we will just keep them all.  

The next step is creating an object of class `disease_expose`, which will allow you to plot and summarize your data very easily. The best way to do this is by calling the helper function to create the object.

```{r eval = FALSE}
de_object <- disease_expose(cleaned_de_data)
```

This pulls up a Shiny gadget and allows you to select your disease column and any exposure/risk factor columns you want to include from your data. After providing your input, an object of class `disease_expose` will be created, barring any errors (most likely that there are columns that are not binary). If Shiny gadgets are not your thing and you would rather manually type in everything you want, you can do that with the constructor below.

```{r}
de_object <- new_disease_expose(cleaned_de_data, disease = 1, exposures = 2:8)
class(de_object)
```

Now that you have an object of class `disease_expose`, you can move on to plotting and summarizing the data, simply by calling `plot` and `summary` and including the object. After the object is created, it is best to not edit it in any way; it should already be in a good format, so no changes are necessary.  

Here is how to plot the data:

```{r fig.width = 7}
#subset used so viewing is better on document
plot(de_object[1:5])
```

As you can see, it produces bar charts for each disease-exposure combination in your data. If you want a little more customization, you can also add arguments that can be added to `geom_bar`. One I like to use is `position = 'dodge'`. Making these plots allows you to visualize how the diseased and non-diseased individuals are distributed across exposure status.

```{r fig.width = 7}
#subset used so viewing is better on document
plot(de_object[1:5], position = 'dodge')
```

In addition to plotting the data, you can also summarize the data like so:

```{r}
summary(de_object)
```

As seen above, you now have many different statistics available for each disease-exposure combination in your data, like odds ratios, incidence in the exposed and unexposed, and confidence intervals for various statistics. Having these statistics available so quickly would be beneficial when seeing if there are associations between any exposures and the disease.  

While the `disease_expose` functions are not overly sophisticated, they do provide some simple and quick tools to analyze disease-exposure data that you might have.  

As seen above, my functions have multiple use cases, as they can help to visualize and summarize patient location history/networks and also visualize and summarize disease-exposure/disease-risk factor data, which could both be useful in a hospital setting.

## Future Work and Plans

I have multiple ideas to improve the package in the future. Those are:

* To create printing methods for the summary methods of my classes. Right now, when you print the output of the `summary` methods, they are not the most pleasing to the eye, as they are just a `data.frame` and a `list`. I would like to create a printing method which makes viewing the results quick and easy to understand, like when viewing the summary output of an `lm` object. I plan to start with the `disease_expose` class, as I think it will be easier to customize output from a `data.frame`. It will most likely resemble the `data.frame` already created, but one idea I had was to highlight the rows with the most significant results (like the highest odds ratio or another statistic). Also, for the `hosp_network` class, the summary output is quite long, so I plan to just select small details that might be the most important when wanting a small overview of the summary. By creating printing methods for both of my summary methods, the summary output will look cleaner and more understandable when printed, which will be an upgrade.
* To write functions to help with survival/time-to-event analysis. These types of analyses could be helpful in a hospital setting, as you might want to study how long a certain hospital-acquired infection takes to infect patients, for example. To do this, I plan to research survival/time-to-event analysis and also collaborate with my internship supervisor. I do not know too much about the topic, so researching it will help me understand how I should build my functions and what different statistics mean. Additionally, collaborating with my internship supervisor will allow me to know what things to include in my functions to make them helpful. I will probably try to do something similar with these functions compared to the others, as I will look to possibly provide plotting and summary functions.
* To provide a more useful plotting function for the `disease_expose` class. Right now, the plotting function plots simple bar charts that show the number of diseased and non-diseased individuals by exposure status. It is a very quick and easy way to visualize disease-exposure data, but it might not be the most useful. Later on, I plan to research common plots that individuals make when studying this type of data, as it might be more useful if I use other types of plots. Since my education does not stretch far into this domain, researching ideas will be the best option, as I can find ideas from experts in the field, rather than relying on my own ideas. If I do not find anything while researching, I will keep the function as is, but if I do find something new, I plan to implement the new idea into a plotting function to make it more useful, while also keeping my simple bar chart plotting function. 
